Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2022-10-24 sports_analytics.Rmd
added new subsection on baseball tools; #31; #32

```{r pkg}
knitr::write_bib(
  c(
    "base",
    "Lahman", 
    "pitchRx", 
    "retrosheet",
    "teamcolors",
    "ggplot2",
    "BradleyTerry2",
    "nflfastR",
    # "nflscrapr", no longer supported, can no longer be installed
    "nflverse",
    "xgboost",
    "chess", 
    "chessR",
    "sportyR",
    "nflplotR"
  ),
  file = "pkgs.bib"
)

@@

```{r}
tibble(
  "American Football" = "nflverse, nflfastR, nflreadr, nfl4th, nflseedR,\n nflplotR, NFLSimulatoR, fflr, ffscrapr, ffsimulator, gsisdecoder, cfbfastR",
  "Association Football (Soccer)" = "worldfootballR, engsoccerdata, socceR, ggsoccer, footballpenaltiesBL, footBayes, itscalledsoccer, FPLdata, EUfootball",
  "Australian Rules Football" = "fitzRoy",
  "Baseball/Softball" = "Lahman, retrosheet, pitchRx, mlbstats, baseballDBR, baseballr, runexp",
  "Basketball" = "BAwiR, AdvancedBasketballStats, uncmbb, BasketballAnalyzeR, NBAloveR, wehoop, hoopR, toRvik",
  "Chess" = "chess, stockfish, bigchess, rchess, chessR",
  "Cricket" = "yorkr, cricketr, cricketdata, howzatR",
  "Esports" = "CSGo, rbedrock, ROpenData, opendotaR, RDota2",
  "GPS Activity Tracking" = "trackeR, trackeRapp, rStrava",
  "Hockey" = "hockeyR, NHLData, nhlapi, nhlscrape, fastRhockey",
  "Swimming" = "SwimmeR",
  "Track and Field" = "combinedevents, JumpeR",
  "Volleyball" = "volleystat"
) |> 
  pivot_longer(everything(),
               names_to = "Sport",
               values_to = "Packages") |>
  transmute(Sport,
            "Number of Packages" = str_count(Packages, ",") + 1,
            "List of Packages" = Packages) |>
  kable(booktabs = TRUE,
        linesep = "",
        caption = "A summary of sport-specific packages available on the Comprehensive R Archive Network (CRAN). \\label{tab:ctv}") |> 
  column_spec(3, width = "7cm")
```

@@

## Graphical tools

@@

>>>>

```{r pkg}
knitr::write_bib(
  c(
    "base",
    "Lahman", 
    "pitchRx", 
    "retrosheet",
    "teamcolors",
    "ggplot2",
    "BradleyTerry2",
    "nflfastR",
    # "nflscrapr", no longer supported, can no longer be installed
    "nflverse",
    "xgboost",
    "chess", 
    "chessR",
    "sportyR",
    "nflplotR",
    "baseballr"
  ),
  file = "pkgs.bib"
)

@@

```{r}
tibble(
  "American Football" = "nflverse, nflfastR, nflreadr, nfl4th, nflseedR,\n nflplotR, NFLSimulatoR, fflr, ffscrapr, ffsimulator, gsisdecoder, cfbfastR",
  "Association Football (Soccer)" = "worldfootballR, engsoccerdata, socceR, ggsoccer, footballpenaltiesBL, footBayes, itscalledsoccer, FPLdata, EUfootball",
  "Australian Rules Football" = "fitzRoy",
  "Baseball/Softball" = "Lahman, retrosheet, pitchRx, mlbstats, baseballDBR, baseballr, runexp",
  "Basketball" = "BAwiR, AdvancedBasketballStats, uncmbb, BasketballAnalyzeR, NBAloveR, wehoop, hoopR, toRvik",
  "Chess" = "chess, stockfish, bigchess, rchess, chessR",
  "Cricket" = "yorkr, cricketr, cricketdata, howzatR",
  "Esports" = "CSGo, rbedrock, ROpenData, opendotaR, RDota2",
  "GPS Activity Tracking" = "trackeR, trackeRapp, rStrava",
  "Hockey" = "hockeyR, NHLData, nhlapi, nhlscrape, fastRhockey",
  "Swimming" = "SwimmeR",
  "Track and Field" = "combinedevents, JumpeR",
  "Volleyball" = "volleystat"
) |> 
  pivot_longer(
    everything(),
    names_to = "Sport",
    values_to = "Packages"
  ) |>
  transmute(
    Sport,
    "Number of Packages" = str_count(Packages, ",") + 1,
    "List of Packages" = Packages
  ) |>
  arrange(desc(`Number of Packages`)) |>
  kable(
    booktabs = TRUE,
    linesep = "",
    caption = "A summary of sport-specific packages available on the Comprehensive R Archive Network (CRAN). While the major North American sports dominate the list, perhaps the fastest-growing collection is for esports. \\label{tab:ctv}"
  ) |> 
  column_spec(3, width = "7cm")
```

@@

## Graphical tools {#sec:graphics}

@@

## Case study in the evolution of tools and research: baseball

As the granularity of baseball data has evolved over time, so too have the statistical methodologies for modeling that data, and the tools for working with it. 
For example, before George Lindsay's work (see Section \ref{sec:ev}), most of the baseball data that was publicly available was seasonal: it showed only season totals for each player. 
Over time, the resolution of baseball data has improved to include play-by-play data, pitch-by-pitch data, and now player tracking data. 
The \pkg{Lahman} package [@R-Lahman] has provided seasonal data for years. 
The \pkg{retrosheet} package [@R-retrosheet] now provides access to the historical play-by-play data available from Retrosheet (this is a comprehensive version of what Lindsay collected for his research). 
The \pkg{pitchRx} package [@R-pitchRx] provides access to pitch-by-pitch data that fueled innovative research into catcher framing [@deshpande2017hierarchical], pitch values [@healey2019bayesian], and pitch classification [@sidle2018using].
The \pkg{baseballr} package [@R-baseballr] now provides access to player tracking data from Statcast.
These data have led to investigations into how defensive shifts affect batting performance [@bouzarth2021swing], as well as how launch angles affect the probability of hitting a home run [@marchi2018analyzing].

As we saw above with chess, the packages in baseball fit together in creative ways. 
In Figure \ref{fig:mlb_standings}, we showed how \pkg{teamcolors} can illuminate data pulled from \pkg{retrosheet} to make an informative data graphic. 
One could just as easily use \pkg{sportyR} to generate a field graphic, and then overlay player tracking data obtained from \pkg{baseballr} to depict defensive shifts.

This, these R packages enable research by making data more easily available. 
Moreover, because R is scriptable, they make it easier to share research that is reproducible. 
[Recent conferences, such as the [Carnegie Mellon Sports Analytics Conference](https://www.stat.cmu.edu/cmsac/conference/2021/), have included a reproducible research competition to foster these efforts.]
